---
title: "Process Tree"
execute:
  echo: false
  warning: false
format: 
  html:
    self-contained: true
editor: visual
---
## Introduction

The purpose of this vignette is to show how to convert nexus trees into input data for the EI coalescent model. 
We will use the maximum clade credibility tree for the 2014 Ebola outbreak in Liberia as our example.

## Read in Tree
We use the phylodyn::summarise_phylo function to produce a list of coalescent and sampling times, as well as the number of lineages sampled at each time point.
We use this list to create data frames which are used as the input into the model.
```{r}
suppressPackageStartupMessages(library(ape))
suppressPackageStartupMessages(library(phylodyn))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggtree))

mcc_tree <- read.nexus(here::here("scripts", "ebola_scripts", "libera_mcc_tree.trees"))
lbr_list <- summarize_phylo(mcc_tree)
```

## Convert Years to Days and Create Data Frames
We now create dataframes of the sampling and coalescent times converted from years (the units of the tree) to days. 
We round to four significant digits.
These data frames are the data needed to run the model
```{r}
sample_frame_lbr <- data.frame(samp_times = lbr_list$samp_times,
                               samp_lin = lbr_list$n_sampled) %>%
  mutate(samp_times_days = samp_times * 365.25,
         samp_times_4dig = round(samp_times_days, 4)) %>%
  group_by(samp_times_4dig) %>%
  summarise(samp_lin = sum(samp_lin))

coal_times_lbr <- data.frame(coal_times = lbr_list$coal_times) %>%
  mutate(coal_times_days = coal_times * 365.25)

```

## Create Date Crosswalk for Visualization 
For visualization purposes it is useful to create a crosswalk between numerical days and dates. 
```{r}
max_reverse_time = max(coal_times_lbr$coal_times_days)
# extract last sample date
fields <- strsplit(mcc_tree$tip.label, "\\|")
# gather all entry 6 into one vector
sample_dates <- sapply(fields, function(x) x[6])
max_date = max(sample_dates)
min_samp_date = min(sample_dates)
reverse_times = seq(0, round(max_reverse_time), by = 1)
reverse_dates = seq(as.Date(max_date), by = "-1 day", length.out = length(reverse_times))
lbr_date_crosswalk <- data.frame(reverse_times = reverse_times,
                                 reverse_dates = reverse_dates,
                                 forward_times = abs(reverse_times - max(reverse_times)))

```
